---
title: "Homework 2"
author: "Arturo Ortiz"
date: "2023-11-06"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Problem #1 #
The first and second component account for about 76% of the variance. Line 5 and 6 contribute the most for comp 1 and line 2 and 3 the most for comp 2
```{r}
crim_corr <- matrix(c(1.000, 0.402, 0.396, 0.301, 0.305, 0.339, 0.340, 0.402, 1.000, 0.618, 0.150, 0.135, 0.206, 0.183, 0.396, 0.618, 1.000, 0.321, 0.289, 0.363, 0.345, 0.301, 0.150, 0.321, 1.000, 0.846, 0.759, 0.661, 0.305, 0.135, 0.289, 0.846, 1.000, 0.797, 0.800, 0.339, 0.206, 0.363, 0.759, 0.797, 1.000, 0.736, 0.340, 0.183, 0.345, 0.661, 0.800, 0.736, 1.000), nrow = 7, byrow = TRUE)

crim_corr.pca <- princomp(covmat = crim_corr)
summary(crim_corr.pca, loadings = TRUE)
```


# Problem #2 #
```{r}
urlremote <- urlRemote <- "https://raw.githubusercontent.com/"
pathGithub <- pathGithub <- "EricBrownTTU/ISQS6350/main/"
filename <- "protein.csv"
protein <- read.csv(paste0(urlremote, pathGithub, filename))
protein <- protein[,1:9]
protein[,1:4]
```
## A ##


```{r}
protein_corr <- cor(protein[,-1])
protein_corr.pca <- princomp(protein[-1], cor = TRUE)
summary(protein_corr.pca, loadings = TRUE)
```



## B ##

The first and second components explains 50.62% and 68.42% of the variance, respectively. 

```{r}
pro_eigen <- eigen(protein_corr)
lambda <- pro_eigen$values
100 * round(cumsum(lambda)/sum(lambda), 4)
```


## C ##

Variables with greater positive loading coefficients such as Eggs and Milk have a positive relationship with Component 1, while those with higher negative loading coefficients such as Cereals and Pulses_nuts_oilseeds have a negative relationship with Component 1. This data assists in determining which variables contribute more or less to Component 1.

Variables with greater positive loading coefficients such as White_Meat and Cereals have a positive relationship with Component 2, while variables with higher negative loading coefficients such as Fish and Starchy_Foods have a negative relationship with Component 2. 



```{r}
load_co <- protein_corr.pca$loadings
round(load_co[,1:2], 3)
```

## D ##

Component 1 and Component 2 in the data show two separate eating habits. Component 1 is linked to a diet high in animal protein and low in starchy foods, where Component 2 is linked to a diet high in white meat, cereals, eggs, and fish but low in starchy foods and legumes/nuts/oilseeds. These interpretations can give insights into the foundation of the dietary data and can be utilized for more study or understanding of the examined population's food preferences and habits.

## E ##
Czechoslovakia and East Germany are 5 and 7 respectively. Czechoslovakia appears to be in the middle of component 1 and component 2. East Germany appears to have a negative correlation for component 2, while having a positive correlation for component 1. 
```{r}
pro.pca <- princomp(protein[,2:9])
biplot(pro.pca, cex = 0.6)
```

# Problem #3 #

```{r}
data("heptathlon", package = "HSAUR2")
hep = heptathlon
hep$hurdles <- max(hep$hurdles) - hep$hurdles
hep$run200m <- max(hep$run200m) - hep$run200m
hep$run800m <- max(hep$run800m) - hep$run800m
head(hep)
```

## A ##
```{r}
heptathlon_scale <- scale(heptathlon[,1:8])
hep.dist <- dist(heptathlon_scale[1:8,])
hep.dist
```

## B ##

mds_result <- cmdscale(hep.dist, k = 2)
eigenvalues <- eigen(crossprod(mds_result))$values
proportions <- eigenvalues / sum(eigenvalues)
target_proportion <- 0.80
cumulative_proportion <- cumsum(proportions)
min_dimensions <- sum(cumulative_proportion < target_proportion) + 1

``` {R} 
cov_hep <- cov(heptathlon[, 1:8])
hep_ev <- eigen(cov_hep)
hep_lambda <- hep_ev$values
round(hep_lambda, 4)
ev <- hep_ev$vectors
round(ev, 4)
cumulative_proportion_cov <- 100 * round(cumsum(hep_lambda) / sum(hep_lambda), 4)
```
## C ##

```{r}
newheptathlon.mds = cmdscale(hep.dist)
plot(newheptathlon.mds, type = "n", main = "Comparison of Athletes", xlab = "1st Component", ylab = "2nd Component")
text(newheptathlon.mds, labels = rownames(heptathlon), cex = 0.4)
```




## D ##

``` {R}
hep.cmd <- cmdscale(hep.dist, k = 2)
plot(hep.cmd, type = "n", main = "Multidimensional Scaling of Heptathlon Data", xlab = "1st Component", ylab = "2nd Component")
text(hep.cmd, labels = rownames(heptathlon), cex = 0.8)
```






